---
title: "data"
format: html
editor: visual
---

## 

\# 2.1

The dataset selected is from the Happy Planet Index (HPI) provided by HotorCool, ranking countries based on their HPI score, using 2021 data with the latest rankings available for 2024. It includes columns HPI rank, Country, ISO, Continent, Population, Life Expectancy, Ladder of Life (Wellbeing), Carbon Footprint, CO2 threshold, GDP per capita, and HPI. Life Expectancy data is sourced from the UN Population Division’s World Population Prospects 2022, updated periodically based on demographic projections. Ladder of Life scores are collected annually by Gallup World Poll through surveys of around 1,000 individuals per country, asking respondents to rate their life satisfaction on a scale of 0 to 10, with results weighted to match national demographics. Carbon Footprint data is taken from the World Inequality Database, providing per capita estimates based on national accounts, and GDP per capita comes from global financial institutions like the World Bank. Some missing values are imputed using adjacent years, creating uncertainty, while survey sampling limitations and time lags in the 2024 rankings (based on 2021 data) may affect accuracy.

The dataset is provided in a excel format. We save the data sheet as a csv file and import it using read_csv.

\# 2.2

```{r}
library(readxl)
library(dplyr)
library(readr)
library(ggplot2)
hpi_2024 <- read_csv(" All countries_20211.csv", skip=8)
```

```{r}
missing_values <- colSums(is.na(hpi_2024))
bar_plot <- ggplot(data.frame(Column = names(missing_values), Missing = missing_values), 
                   aes(x = Column, y = Missing, fill = Column)) +
  geom_bar(stat = "identity", color = "black") +
  labs(title = "Count of Missing Values per Column", x = "Columns", y = "Count of Missing Values") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1)) + # Tilt x-axis labels
  scale_fill_brewer(palette = "Set3")
bar_plot
```

The data does not exhibit any obvious missing values. However, as mentioned in the HPI 2024 methodology report, some missing values may have been imputed using data from adjacent years.

```{r}
hpi_2024$IsMissing <- ifelse(hpi_2024$`Ladder of life (Wellbeing) (0-10)` == 0, "0", "Other Values")

missing_plot <- ggplot(hpi_2024, aes(x = IsMissing, fill = IsMissing)) + geom_bar(color = "black") + scale_fill_manual(values = c("Missing (0)" = "pink", "Other Values" = "lightblue")) + labs(title = "Distribution of Missing (0) vs Other Values", x = "Value Type", y = "Count") + theme_minimal()

print(missing_plot)
```

However, there are 0 values in the HPI scores and the Ladder of Life columns, which may indicate missing data in the Ladder of Life scores. Based on the calculation methodology for the HPI score, a 0 in the HPI score might be directly influenced by a 0 in the Ladder of Life column. Therefore, we analyze the number of 0 values in the Ladder of Life column to observe the potential missing values. We can see that there are only two 0 scores for Ladder of Life columns, which are for Sudan and Burundi.
