---
title: "data"
format: html
editor: visual
---

## Technical Description

The data set selected is from the Happy Planet Index (HPI) provided by HotorCool. It contains rankings for countries from 2006-2021 based on data from the corresponding years. For year, the data set includes columns HPI rank, Country, ISO, Continent, Population, Life Expectancy, Ladder of Life (Wellbeing), Carbon Footprint, CO2 threshold, GDP per capita, and HPI. For the latest 2024 data set, Life Expectancy data is sourced from the UN Population Division’s World Population Prospects 2022, updated periodically based on demographic projections. Ladder of Life scores are collected annually by Gallup World Poll through surveys of around 1,000 individuals per country, asking respondents to rate their life satisfaction on a scale of 0 to 10, with results weighted to match national demographics. Carbon Footprint data is taken from the World Inequality Database, providing per capita estimates based on national accounts, and GDP per capita comes from global financial institutions like the World Bank. Some missing values are imputed using adjacent years, creating uncertainty, while survey sampling limitations and time lags in the 2024 rankings may affect accuracy.

The data set is provided in a excel format. We save the data sheet for each year as a csv file and integrate the data sets from 2006 to 2021 to one in R.

## Missing value analysis

```{r, message = FALSE, warning = FALSE}
library(readxl)
library(dplyr)
library(readr)
library(ggplot2)
<<<<<<< HEAD
hpi_2024 <- read_csv("All countries_20211.csv", skip=8)
```

```{r, message = FALSE, warning = FALSE}
missing_values <- colSums(is.na(hpi_2024))
=======
hpi <- read_csv("data/hpi.csv")
```

```{r}
hpi <- hpi %>% select(-`...5`, -`...14`, -`...1`)
missing_values <- colSums(is.na(hpi))
>>>>>>> 58df64aabbd27501d8e707824f4591e9b95a5007
bar_plot <- ggplot(data.frame(Column = names(missing_values), Missing = missing_values), 
                   aes(x = Column, y = Missing, fill = Column)) +
  geom_bar(stat = "identity", color = "black") +
  labs(title = "Count of Missing Values per Column", x = "Columns", y = "Count of Missing Values") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1)) + 
  scale_fill_brewer(palette = "Set3")
bar_plot
```

The data does not exhibit any obvious missing values after we remove the three columns that are duplicate with country labels. However, as mentioned in the HPI 2024 methodology report, some missing values may have been imputed using data from adjacent years.

<<<<<<< HEAD
```{r, message = FALSE, warning = FALSE}
hpi_2024$IsMissing <- ifelse(hpi_2024$`Ladder of life (Wellbeing) (0-10)` == 0, "0", "Other Values")
=======
```{r}
hpi$IsMissing <- ifelse(hpi$`Ladder of life (Wellbeing) (0-10)` == 0, "0", "Other Values")
>>>>>>> 58df64aabbd27501d8e707824f4591e9b95a5007

missing_plot <- ggplot(hpi, aes(x = IsMissing, fill = IsMissing)) + geom_bar(color = "black") + scale_fill_manual(values = c("0" = "pink", "Other Values" = "lightblue")) + labs(title = "Distribution of Value 0 vs Other Values", x = "Value Type", y = "Count") + theme_minimal()

print(missing_plot)
```

However, there are 0 values in the HPI scores and the Ladder of Life columns, which likely indicate missing data in the Ladder of Life scores. Given the calculation methodology for the HPI score, a 0 in the HPI score is caused by a 0 in the Ladder of Life column. To better understand the extent of missing data, we analyzed the number of 0 values in the Ladder of Life column and found that there are approximately 250 occurrences across the years. For the analysis, countries with 0 values may be excluded as they do not contribute meaningful information to the results.
