# Results

```{r, message = FALSE, warning = FALSE}
library(readxl)
library(dplyr)
library(readr)
library(ggplot2)
library(tidyr)
library(ggrepel)
library(tidyverse)
library(reshape2)
hpi_2006 <- read_csv("hpi_2006.csv")
hpi_2011 <- read_csv("hpi_2011.csv")
hpi_2016 <- read_csv("hpi_2016.csv")
hpi_2021 <- read_csv("hpi_2021.csv")
hpi_2024 <- read_csv("All countries_20211.csv", skip=8, show_col_types = FALSE)
past_HPI <- read_csv("Historical_HPI.csv", show_col_types = FALSE)
past_HPI[is.na(past_HPI)] <- 0
```

## HPI through the years
Let's investigate how HPI varies across years. The plot below shows the HPI for top 20 countries from 2006 to 2021. It is noticeable that Costa Rica has the highest mean HPI, and Sweden has the highest HPI in 2021. Overall, Costa Rica shows obviously high happiness. 

```{r, message = FALSE, warning = FALSE}
past_HPI$Mean_HPI <- rowMeans(past_HPI[, 2:17], na.rm = TRUE)

past_HPI_long <- past_HPI %>%
  pivot_longer(cols = `2006`:`2021`, names_to = "Year", values_to = "HPI") %>%
  mutate(Year = as.numeric(gsub("HPI_", "", Year))) # Extract year from column names

top_countries <- past_HPI %>%
  arrange(desc(Mean_HPI)) %>%
  head(20) %>%
  arrange(Mean_HPI)

past_HPI_long <- past_HPI_long %>%
  filter(Country %in% top_countries$Country)

# Create the plot
ggplot(data = past_HPI_long, aes(x = HPI, y = reorder(Country, Mean_HPI))) +
  # Historic scores (keep light gray as before)
  geom_point(color = "gray", size = 1.5, alpha = 0.5) +
  # Mean score (solid yellow, smaller size)
  geom_point(data = top_countries, aes(x = Mean_HPI, y = reorder(Country, Mean_HPI)),
             color = "#33a02c", size = 3, shape = 16) +
  # 2021 score (solid blue, smaller size)
  geom_point(data = filter(past_HPI_long, Year == 2021),
             aes(x = HPI, y = Country), color = "#4682B4", size = 3, shape = 16) +
  # Add dashed horizontal lines for countries
  geom_hline(yintercept = seq_along(top_countries$Country), color = "gray40", linetype = "dashed", size = 0.2) +
  # Titles and labels
  labs(title = "Happiness Index through the Years",
       subtitle = "The happiness levels are largely consistnet across the countries.\nCosta Rica's mean score is the highest, while Sweden's 2021 score is the highest.",
       x = "Happiness Index", y = NULL) +
  # Theme adjustments
  theme_minimal(base_family = "sans") +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(hjust = 1, color = "gray40"),
    axis.title.x = element_text(hjust = 0, color = "gray40"),
    plot.title = element_text(size = 17, face = "bold"),
    plot.subtitle = element_text(size = 12, color = "gray40")
  )+
  annotate("text", x = max(past_HPI_long$HPI, na.rm = TRUE), y = 1, 
           label = "2021 score", hjust = 1.5, size = 4, color = "#4682B4") +
  annotate("text", x = max(top_countries$Mean_HPI, na.rm = TRUE), y = 2, 
           label = "Mean score", hjust = 1.5, size = 4, color = "#33a02c")
```

## Overall trend from 2006 to 2021 for all countries

To further learn the pattern of HPI from 2006 to 2021, we take all countries in account instead of only the top 20 ones. We pick year 2006, 2011, 2016, and 2021 for comparison.

```{r}
hpi_2006_hist <- hpi_2006 %>% 
  mutate(Year = 2006) %>% 
  filter(HPI > 0)  

hpi_2011_hist <- hpi_2011 %>% 
  mutate(Year = 2011) %>% 
  filter(HPI > 0) 

hpi_2016_hist <- hpi_2016 %>% 
  mutate(Year = 2016) %>% 
  filter(HPI > 0) 

hpi_2021_hist <- hpi_2021 %>% 
  mutate(Year = 2021) %>% 
  filter(HPI > 0)  


combined_data <- bind_rows(hpi_2006_hist, hpi_2011_hist, hpi_2016_hist, hpi_2021_hist)
ggplot(combined_data, aes(x = HPI)) +
  geom_histogram(binwidth = 5, fill = "#4682B4", color = "white", alpha = 0.8, boundary = 0) +
  facet_wrap(~ Year, ncol = 2) +  
  theme_minimal(base_size = 12) +
  
  theme(
    panel.grid.major = element_line(color = "gray85"),
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 10, color = "black"),
    axis.title = element_text(size = 12, color = "black"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  ) +
  scale_x_continuous(
    breaks = seq(0, 60, by = 10), 
  ) +
  labs(
    title = "Distribution of Happy Planet Index (HPI) by Year",
    x = "HPI Score",
    y = "Count"
  )

```

This plot shows the distribution of HPI scores for year 2006, 2011, 2016, 2021. The year 2006 marks the starting point of the survey, while 2021 represents the most recent data. Rows with HPI scores of 0, which indicate missing values, were excluded from the dataset. The distribution reveals notable variability across the years.
In 2006, the distribution appears slightly right-skewed, with a noticeable peak around 30-40. In 2011, the distribution becomes slightly symmetrical, with most scores concentrated between 30 and 50. In 2006, the distribution shifts slightly upward, with a noticeable concentration around 35-50. In 2021, the distribution shifts downward, with most scores falling between 25 and 50. This plot highlights how global well-being has evolved over time.


## Continental view

Grouped by continents, let's see how each of them perform! To clearly visualize the HPI by continents, we regroup 8 of them to 4 in total. 

```{r, message = FALSE, warning = FALSE}
hpi_2024 <- hpi_2024 %>%
  mutate(Region_Group = case_when(
    Continent %in% c(1, 2) ~ "Americas",                  # Latin America, N. America & Oceania
    Continent %in% c(3, 7) ~ "Europe & Central Asia",     # Western Europe, Eastern Europe & Central Asia
    Continent %in% c(4, 6, 8) ~ "Asia & Middle East",     # Middle East, South Asia, East Asia
    Continent == 5 ~ "Africa",                            # Africa                                
  ))

ggplot(hpi_2024, aes(
  x = `Ladder of life (Wellbeing) (0-10)`,
  y = HPI,                                 
  color = Region_Group                     
)) +
  geom_point(size = 3, alpha = 0.8) +
  labs(
    title = "Happiness Index and Life Expectancy by Region",
    subtitle = "Relationship between HPI and Wellbeing across regions",
    x = "Ladder of life (Wellbeing)",
    y = "Happiness Index",
    color = "Region Group"
  ) +
  scale_color_manual(
    values = c(
      "Americas" = "#4682B4",
      "Europe & Central Asia" = "#33a02c",
      "Asia & Middle East" = "#CC6666",
      "Africa" = "#D97D0D"
    )
  ) +
  # Customize the theme
  theme_minimal(base_family = "sans") +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "#fbfbfb"),
    plot.background = element_rect(fill = "#fbfbfb"),
    plot.title = element_text(size = 17, face = "bold", color = "#323232"),
    plot.subtitle = element_text(size = 12, face = "italic", color = "gray40"),
    axis.title.x = element_text(size = 10, face = "bold", color = "#323232"),  # Removed `family`
    axis.title.y = element_text(size = 10, face = "bold", color = "#323232"),
    legend.position = "right",
    legend.background = element_blank(),
    legend.box.background = element_rect(fill = "transparent", color = NA)
  )
```
We can see that Americas and Europe & Central Asia tend to have high HPI and Wellbeing; Asia & Middle East has a wide spread, which shows large variation; and Africa shows obviously low HPI and Wellbeing. From this plot, we can also recognize that there is a strong positive relationship. 

## Continental Variance

To better understanding the differences between continents, we plot boxplots for all of the 8 continent.
```{r, message = FALSE, warning = FALSE}
library(RColorBrewer)
continent_mapping <- c(
  "Latin America",
  "North America & Oceania",
  "Western Europe",
  "Middle East & North Africa",
  "Africa",
  "South Asia",
  "Eastern Europe & Central Asia",
  "East Asia"
)
hpi_2021$ContinentName <- continent_mapping[hpi_2021$Continent]
qualitative_palette <- brewer.pal(n = length(unique(hpi_2021$ContinentName)), "Set3")

ggplot(hpi_2021, aes(x = HPI, y = ContinentName, fill = ContinentName)) +
  geom_boxplot(width = 0.6, alpha = 0.8) + 
  scale_fill_manual(values = qualitative_palette) + 
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(size = 10, color = "black"),
    axis.text.x = element_text(size = 8, color = "black"),
    legend.position = "none",
    panel.border = element_blank(),
    axis.line = element_blank()
  ) +
  labs(
    title = "Continental Variance In 2021",
    x = "HPI Score",
    y = ""
  )
```
This boxplot illustrates the variance of HPI scores across 8 continents with 2021 ranking. Western Europe, Latin America, and North America & Oceania show higher overall scores and tighter distributions, suggesting consistent well-being. In contrast, Africa and the Middle East & North Africa exhibit wider variability, indicating more significant disparities in well-being within these regions. This analysis emphasizes the importance of regional and local context when addressing global well-being challenges.

## Differences between happy and unhappy countries

We define happy countries if their HPI is higher than the average, otherwise they are unhappy countries. Five variables are included here: population, life expectancy, wellbeing, GDP per capita, and carbon foorprint. 
```{r, message = FALSE, warning = FALSE}
library(gridExtra)
library(grid) 
library(patchwork)
#Split Data Based on Average HPI
hpi_avg <- mean(hpi_2024$HPI, na.rm = TRUE)  # Calculate average HPI
hpi_2024 <- hpi_2024 %>%
  mutate(Lower_Happy = ifelse(HPI < hpi_avg, "Unhappy", "Happy"))  # Add split column

variables <- c(
  "Population (thousands)",
  "Life Expectancy (years)",
  "Ladder of life (Wellbeing) (0-10)",
  "GDP per capita ($)",
  "Carbon Footprint (tCO2e)"
)
titles <- c(
  "Population Distribution",
  "Life Expectancy Distribution",
  "Wellbeing Distribution",
  "GDP per Capita Distribution",
  "Carbon Footprint (tCO2e)"
)

plot_list <- lapply(seq_along(variables), function(i) {
  var <- variables[i]  
  ggplot(hpi_2024, aes_string(x = paste0("`", var, "`"), fill = "Lower_Happy")) +
    geom_density(alpha = 0.7, color = "black") +
    scale_fill_manual(values = c("#4682B4", "#CC6666")) +  # Customize colors
    labs(x = var, y = NULL, fill = "Group", title = titles[i]) +  # Add plot title
    theme_minimal() +
    theme(
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank(),
      panel.grid.major.x = element_line(color = "gray", linetype = "dotted"),
      panel.grid.minor = element_blank(),
      axis.title.x = element_text(size = 10, face = "bold", family = "monospace"),
      plot.title = element_text(size = 10, face = "bold", hjust = 0.5),  
      legend.position = "bottom",
      legend.title = element_text(size = 8, face = "bold"), 
      legend.text = element_text(size = 6),  
      legend.key.size = unit(0.5, "cm") 
    )
})

combined_plot <- wrap_plots(plot_list, ncol = 3) +
  plot_layout(guides = "collect") +  # Collect a single shared legend
  plot_annotation(
    title = "Differences Between Happy & Unhappy Countries",
    subtitle = "There are large differences, with Life Expectancy and Wellbeing.",
    theme = theme(
      plot.title = element_text(size = 15, face = "bold", hjust = 0.5, color = "#323232"),
      plot.subtitle = element_text(size = 12, face = "italic", hjust = 0.5, color = "gray"),
      legend.position = "bottom"  
    )
  )

# Print the final combined plot
print(combined_plot)

```
From the plot, happy countries show remarkable high life expectancy and wellbeing, which is consistent to the trend we explored on the previous plot. Unhappy countries have longer right tails for carbon footprint, which is reasonable since extremely high carbon emission will reduce their quality of life. A great proportion of unhappy countries have extremely low GDP per capita. It is surprising that many unhappy countries have low carbon footprint, but at the same time, some of them have very high carbon footprint. Low carbon foot print may be due to lack of technology in industries, so they do not have many machines working. 

## Linear relatinship between HPI and Ladder of Life

Since the scatter plot for continents already show the relationship between HPI and wellbeing, now we wan to investigate the relationship bewteen HPI and life expectancy. 
```{r, message = FALSE, warning = FALSE}

hpi_2024 <- hpi_2024 %>%
  rename(
    HPI = `HPI`,  # Keep HPI column name
    Life_Expectancy = `Life Expectancy (years)` 
  ) %>%
  mutate(
    HPI = as.numeric(HPI), 
    Life_Expectancy = as.numeric(Life_Expectancy)
  )

# Remove rows with missing data
hpi_clean <- hpi_2024 %>%
  filter(!is.na(HPI), !is.na(Life_Expectancy))

model <- lm(HPI ~ Life_Expectancy, data = hpi_clean)

ggplot(hpi_clean, aes(x = Life_Expectancy, y = HPI)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(
    title = "Relationship Between HPI and Life Expectancy",
    x = "Life Expectancy (years)",
    y = "HPI"
  ) +
  theme_minimal()
```
The linear regression analysis indicates a significant positive relationship between Life Expectancy and Happiness Index (HPI). The regression coefficient for Life Expectancy is 0.812, meaning that for every additional year of life expectancy, the HPI is expected to increase by approximately 0.81 points. The results highlight that Life Expectancy is a significant contributor of HPI.

## See how random countries vary across years

Here 8 randomly picked countries are plotted for their HPI in 2006 and 2021. Countries with missing value for HPI are removed for a better visualization. We want to see what pattern we can learn.
```{r, message = FALSE, warning = FALSE}
# Randomly sample 8 countries
set.seed(809)
sample <- past_HPI %>%
  filter(`2006` != 0, `2021` != 0) %>%  # Exclude countries with HPI of 0 for 2006 or 2021
  slice_sample(n = 8) %>%
  select(Country, `2006`, `2021`)


countries_2006 <- sample %>%
  group_by(`2006`) %>%
  summarize(Country = paste(Country, collapse = ", "))

countries_2021 <- sample %>%
  group_by(`2021`) %>%
  summarize(Country = paste(Country, collapse = ", "))

sample_long <- sample %>%
  pivot_longer(cols = c(`2006`, `2021`), names_to = "Year", values_to = "HPI") %>%
  mutate(Year = as.numeric(Year))

options(repr.plot.width = 15, repr.plot.height = 8) 
ggplot() +
  geom_vline(xintercept = c(1, 2), color = "black", linetype = "dotted", size = 0.5) +
  geom_segment(data = sample, aes(
    x = 1, xend = 2, y = `2006`, yend = `2021`, color = ifelse(`2021` > `2006`, "#1f78b4", "#e31a1c")
  ), size = 1) +
  
  geom_point(data = sample, aes(
    x = 1, y = `2006`, color = ifelse(`2021` > `2006`, "#1f78b4", "#e31a1c")
  ), size = 4) +

  geom_point(data = sample, aes(
    x = 2, y = `2021`, color = ifelse(`2021` > `2006`, "#1f78b4", "#e31a1c")
  ), size = 4) +
 
  geom_text(data = countries_2006, aes(
    x = 1 - 0.05, y = `2006`, label = Country
  ), hjust = "right", size = 2.5) +
  
  geom_text(data = countries_2021, aes(
    x = 2 + 0.05, y = `2021`, label = Country
  ), hjust = "left", size = 2.5) +
 
  labs(
    title = "Happiness Index Scores: 2006 vs 2021",
    subtitle = "Random sample of countries showing HPI changes between 2006 and 2021",
    x = NULL,
    y = "HPI"  
  ) +

  scale_color_identity() +
  scale_x_continuous(limits = c(0.5, 2.5),breaks = c(1, 2), labels = c("2006", "2021")) +
  scale_y_continuous(limits = c(min(sample_long$HPI) , max(sample_long$HPI))) + 
  theme_minimal() +
  theme(
    plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 14, hjust = 0.5),
    axis.text.y = element_text(size = 12),
    axis.text.x = element_text(size = 12, face = "bold"),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    legend.position = "none"
  )

```
Countries like Nicaragua show significant improvement, with the highest increase in HPI, indicating progress in happiness. Nepal, and Sri Lanka remain relatively stable, indicating limited change. On the other hand, Rwanda and Kazakhstan show noticeable declines in HPI, reflecting potential challenges in maintaining life satisfaction. 

## What's biggest movers?

```{r, message = FALSE, warning = FALSE}
# Filter the top 20 biggest movers
top_20 <- past_HPI %>%
  filter(`2006` != 0, `2021` != 0) %>%
  mutate(Change = abs(`2021` - `2006`)) %>%
  arrange(desc(Change)) %>%
  slice(1:20) 

top_20 <- top_20 %>%
  mutate(Highlight = ifelse(row_number() <= 3, "Top Increase", 
                            ifelse(row_number() > (n() - 3), "Top Decrease", "Neutral")))


plot_data <- top_20 %>%
  pivot_longer(cols = c(`2006`, `2021`), names_to = "Year", values_to = "HPI") %>%
  mutate(Year = as.numeric(Year))

low_c <- "#CC6666"
high_c <- "#33a02c" 
neutral_c <- "#e7e9e7"


ggplot(top_20, aes(x = reorder(Country, Change), y = Change, fill = Highlight)) +
  geom_bar(stat = "identity", alpha = 0.7) +
  scale_fill_manual(values = c("Top Increase" = high_c, "Top Decrease" = low_c, "Neutral" = neutral_c)) +
  theme_minimal(base_family = "sans") +
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8),
    axis.text.y = element_text(size = 8),
    panel.grid = element_blank(),
    panel.background = element_rect(fill = "#fafafa", color = NA),
    plot.background = element_rect(fill = "#fafafa", color = NA),
    legend.position = "none"
  ) +
  labs(
    title = "The Biggest Movers",
    subtitle = "These countries have had the largest change in HPI between 2006 and 2021"
   
  )

```
This bar chart displays the top 20 countries with the largest changes in their Happy Planet Index (HPI) between 2006 and 2021, with green bars representing increases and red bars representing decreases. The countries with the most significant positive changes include Lebanon, Nicaragua, and India, while the most substantial negative changes are observed in Latvia, Benin, and Uzbekistan. From this plot, we can clearly see that HPI movements can vary a lot for different countries.

## Other factors

```{r, message = FALSE, warning = FALSE}

hpi_2024 <- hpi_2024 %>%
  mutate(
    Above_CO2_Threshold = ifelse(
      `Carbon Footprint (tCO2e)` > `CO2 threshold for year  (tCO2e)`, 1, 0
    )
  )


heatmap_data <- hpi_2024 %>%
  select(
    HPI,
    Population = `Population (thousands)`,
    Life_Expectancy,
    Ladder_of_Life = `Ladder of life (Wellbeing) (0-10)`,
    GDP_per_capita = `GDP per capita ($)`,
    Above_CO2_Threshold
  )

# Ensure numeric columns for correlation (convert boolean if needed)
heatmap_data$Above_CO2_Threshold <- as.numeric(heatmap_data$Above_CO2_Threshold)

# Compute the correlation matrix
cor_matrix <- cor(heatmap_data, use = "complete.obs")

# Melt the correlation matrix for ggplot
melted_cor_matrix <- melt(cor_matrix)

# Create the heatmap
ggplot(data = melted_cor_matrix, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(value, 2)), color = "black", size = 3) +  # Add correlation values
  scale_fill_gradient(low = "white", high = "#8C2B3D", name = "Correlation") +
  labs(
    title = "Heatmap of HPI and Selected Variables",
    x = "Variables",
    y = "Variables"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  )

```
The heatmap highlights that subjective well-being (Ladder of Life) and Life Expectancy are the strongest predictors of HPI, underscoring the importance of health and perceived quality of life in fostering happiness. Economic factors, while relevant, play a secondary role, and ecological sustainability (Above CO2 Threshold) moderately supports happiness. 

## PCA Analysis of Wellbeing, Life Expectancy, and the Happy Planet Index

Since we already kniw that ladder of life and life expectancy are the top two factors contributing to higher HPI, now we will explore further. 
```{r, message = FALSE, warning = FALSE}
pca_data <- hpi_2024 %>%
  select(
    Ladder_of_Life = `Ladder of life (Wellbeing) (0-10)`,
    Life_Expectancy,
    HPI
  ) %>%
  drop_na()  # Remove rows with missing values

# Standardize the data
pca_scaled <- scale(pca_data)

# Perform PCA
pca_result <- prcomp(pca_scaled, center = TRUE, scale. = TRUE)

# Extract PCA scores (for points)
pca_scores <- as.data.frame(pca_result$x)
pca_scores <- pca_scores %>%
  mutate(HPI = pca_data$HPI)  # Add HPI back for context in visualization

# Extract PCA loadings (for vectors)
pca_loadings <- as.data.frame(pca_result$rotation)
pca_loadings <- pca_loadings %>%
  rownames_to_column(var = "Variable")

# Visualize PCA with points and vectors
ggplot() +
  # Plot PCA scores (observations)
  geom_point(data = pca_scores, aes(x = PC1, y = PC2, color = HPI), size = 4, alpha = 0.7) +
  scale_color_gradient(low = "blue", high = "red", name = "HPI") +
  # Plot PCA loadings (vectors)
  geom_segment(data = pca_loadings, aes(
    x = 0, y = 0, xend = PC1 * 5, yend = PC2 * 5  # Scale vectors for visibility
  ), arrow = arrow(length = unit(0.2, "cm")), color = "black") +
  geom_text_repel(data = pca_loadings, aes(
    x = PC1 * 5, y = PC2 * 5, label = Variable
  ), size = 5, color = "black") +
  # Add titles and labels
  labs(
    title = "PCA Biplot: Ladder of Life and Life Expectancy",
    x = "PC1 (Ladder of Life)",
    y = "PC2 (Life Expectancy)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 12, face = "bold")
  )


```
The PCA biplot illustrates the relationship between the "Ladder of Life" (PC1), "Life Expectancy" (PC2), and the Happy Planet Index (HPI). PC1 explains the variability primarily associated with subjective life wellbeing, while PC2 explains variability related to life expectancy. The arrows indicate that "Ladder of Life" and "Life Expectancy" contribute significantly to these components, with "HPI" showing a positive correlation with both. Countries with higher HPI values (red dots) tend to cluster around higher values of PC1 and PC2, suggesting that life satisfaction and life expectancy positively contribute to the HPI. Conversely, countries with low HPI values (blue dots) show lower PC1 and PC2 scores, reflecting poorer outcomes in these dimensions.



